---
title: "Colts Game 1 (2025): Data Pull, Integrity Checks, and EDA"
author: "Grant Starnes"
date: "2025-09-20"
output: html_document
---

### First, we want to load the necessary libraries from the original NFLReadrPhaseI document

```{r}
library(nflreadr)    # NFL data
library(tidyverse)   # dplyr, ggplot2, etc.
library(lubridate)   # Dates & times
library(janitor)     # clean_names
library(scales)      # Pretty labels
library(gt)          # Nice tables
library(ggrepel)     # Text labels on charts
library(zoo)         # Rolling calculations
```

### Next, we want to set our parameters. We want the team to be the Colts and the season to be the 2025 season

```{r}
team   <- "IND"    # Colts
season <- 2025     # Season
```

### In this section, we want to access the data and locate the Colts' first regular-season game in which they played the Miami Dolphins

```{r}
sched <- load_schedules(seasons = season) %>% clean_names()

colts_g1 <- sched %>%
  filter(game_type == "REG",
         season == !!season,
         home_team == !!team | away_team == !!team) %>%
  arrange(gameday, week) %>%
  slice(1)

stopifnot(nrow(colts_g1) == 1)

game_id <- colts_g1$game_id
week    <- colts_g1$week
gameday <- colts_g1$gameday
home_tm <- colts_g1$home_team
away_tm <- colts_g1$away_team
opp     <- ifelse(home_tm == team, away_tm, home_tm)

glue::glue("Found Colts Week {week} ({gameday}): {away_tm} @ {home_tm} | game_id = {game_id}")
```

### Following that, we then want to pull the essential data, which is the play-by-play data from Game 1, the rosters for both teams, and the officials for the game

```{r}
pbp_g1 <- load_pbp(seasons = season) %>%
  clean_names() %>%
  filter(game_id == !!game_id)

rw_g1 <- load_rosters_weekly(seasons = season) %>%
  clean_names() %>%
  filter(week == !!week, team %in% c(team, opp))

officials_g1 <- load_officials(seasons = season) %>%
  clean_names() %>%
  filter(game_id == !!game_id)

players_master <- load_players() %>% clean_names()
```

### Moving forward, in this block we'd like to set the dimensions of the associated datasets. In this case, the datasets for play-by-play, rosters, and officials

```{r}
cat("PBP rows/cols:", nrow(pbp_g1), ncol(pbp_g1), "\n")
cat("Weekly roster rows/cols:", nrow(rw_g1), ncol(rw_g1), "\n")
cat("Officials rows/cols:", nrow(officials_g1), ncol(officials_g1), "\n\n")
```

### Following setting the dimensions, we now want to ensure there are no duplicate plays in the play-by-play data, and ensure all plays are unique

```{r}
dup_plays <- pbp_g1 %>%
  count(game_id, play_id) %>%
  filter(n > 1)

if (nrow(dup_plays) > 0) {
  warning("Duplicate plays detected in (game_id, play_id). Inspect `dup_plays`.")
} else {
  cat("Key check passed: (game_id, play_id) appear unique for this game.\n")
}

# For weekly rosters, (team, gsis_id) often repeats by position slots; check basic duplication
dup_roster <- rw_g1 %>%
  count(team, gsis_id, week) %>%
  filter(n > 1)
if (nrow(dup_roster) > 0) {
  message("Note: roster has repeated (team, gsis_id, week) rows—can be normal when multiple rows exist per role/slot.")
}
```

### Next, we want to check for referential integrity between the roster and master player data

```{r}
if ("gsis_id" %in% names(rw_g1) && "gsis_id" %in% names(players_master)) {
  roster_master_match <- rw_g1 %>%
    mutate(in_master = gsis_id %in% players_master$gsis_id) %>%
    summarize(match_rate = mean(in_master, na.rm = TRUE))
  cat("Player master match rate (weekly roster -> master by gsis_id):",
      round(roster_master_match$match_rate*100, 1), "%\n")
} else {
  message("gsis_id not present to check roster->player master referential integrity.")
}
```

### Here, we want to provide a summary of missingness in the play-by-play data

```{r}
na_rate <- function(x) mean(is.na(x))
na_summary <- pbp_g1 %>%
  summarize(
    na_epa          = na_rate(epa),
    na_success      = na_rate(success),
    na_posteam      = na_rate(posteam),
    na_defteam      = na_rate(defteam),
    na_yardline_100 = na_rate(yardline_100),
    na_down         = na_rate(down),
    na_ydstogo      = na_rate(ydstogo),
    na_play_type    = na_rate(play_type)
  )
cat("\n--- MISSINGNESS (PBP) ---\n"); print(na_summary)
```

### Here, we want to make sure the the yardages for the play-by-play data is within the correct range, 0-100 yards, the dimensions of a standard football field.

```{r}
# Yardline must be 0-100
bad_yardline <- pbp_g1 %>% filter(!is.na(yardline_100) & (yardline_100 < 0 | yardline_100 > 100))
if (nrow(bad_yardline) > 0) warning("Out-of-range yardline_100 values observed.")
```

### We also want to make sure the downs are between 1-4, the standard measurement of a drive regarding the down in football

```{r}
# Down must be 1-4
bad_down <- pbp_g1 %>% filter(!is.na(down) & !(down %in% 1:4))
if (nrow(bad_down) > 0) warning("Unexpected down values outside 1–4.")
```

### Bad Yards to Go

```{r}
bad_ydstogo <- pbp_g1 %>% filter(!is.na(ydstogo) & ydstogo <= 0)
if (nrow(bad_ydstogo) > 0) message("Non-positive ydstogo rows exist (could be goal-to-go anomalies or data glitches).")
```

### Here, we want to make sure the game_id is set to 1 for the first game of the season

```{r}
if (length(unique(pbp_g1$game_id)) != 1) warning("Multiple game_ids in pbp_g1; filter may be off.")
```

### Posteam/Defteam

```{r}
valid_teams <- c(team, opp)
weird_teams <- pbp_g1 %>%
  filter(!is.na(posteam) & !(posteam %in% valid_teams)) %>%
  distinct(posteam)
if (nrow(weird_teams) > 0) warning("Found posteam not matching Colts or opponent: check `weird_teams`.")
```

### Team Abbreviations

```{r}
cat("\n--- TEAM ABBREVS IN PBP ---\n")
print(sort(unique(na.omit(pbp_g1$posteam))))
print(sort(unique(na.omit(pbp_g1$defteam))))
```

### In this section, we want to set each way someone can score to an associated score value

```{r}
pbp_g1 <- pbp_g1 %>%
  mutate(
    points_scored = case_when(
      touchdown == 1 ~ 6,                               # Touchdowns
      field_goal_result == "made" ~ 3,                  # Field goals
      safety == 1 ~ 2,                                  # Safeties
      extra_point_result == "good" ~ 1,                 # PAT kicks
      two_point_attempt == 1 & two_point_conv_result == "success" ~ 2, # 2PT conversions
      TRUE ~ 0
    )
  )
```

### Next, we want to find out the scoring in regards to quarter

```{r}
score_by_q <- pbp_g1 %>%
  filter(!is.na(qtr), qtr %in% 1:4) %>%
  group_by(qtr) %>%
  summarize(
    ind_points = sum(if_else(posteam == "IND", points_scored, 0), na.rm = TRUE),
    opp_points = sum(if_else(posteam == opp,   points_scored, 0), na.rm = TRUE),
    plays = n(),
    .groups = "drop"
  )
cat("\n--- SCORE BY QUARTER (derived from PBP points events) ---\n")
print(score_by_q)

tempo <- pbp_g1 %>%
  filter(!is.na(posteam), !is.na(game_seconds_remaining)) %>%
  arrange(game_seconds_remaining) %>%
  group_by(posteam) %>%
  mutate(sec_between = lag(game_seconds_remaining) - game_seconds_remaining) %>%
  summarize(sec_per_play = median(sec_between, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(sec_per_play))
cat("\n--- TEMPO (median sec/play) ---\n"); print(tempo)
```

### Here, we will generate a plot associated with scoring per quarter

```{r}
ggplot(score_by_q, aes(x = factor(qtr))) +
  geom_col(aes(y = ind_points, fill = "Colts"), position = "dodge") +
  geom_col(aes(y = opp_points, fill = "Opponent"), position = "dodge") +
  labs(title = "Scoring by Quarter", x = "Quarter", y = "Points", fill = "Team") +
  theme_minimal()
```

### Here, we want to determine the offensive play mix, which can be one of the following: Run, Pass, Clock (QB Kneel, QB Spike), or Other for plays not included in these categories. We can also determine its efficiency and success rate

```{r}
off_mix <- pbp_g1 %>%
  filter(!is.na(posteam)) %>%
  mutate(play_family = case_when(
    play_type %in% c("run") ~ "Run",
    play_type %in% c("pass") ~ "Pass",
    play_type %in% c("qb_kneel","qb_spike") ~ "Clock",
    play_type %in% c("no_play","timeout") ~ "Other",
    TRUE ~ "Other"
  )) %>%
  group_by(posteam, play_family) %>%
  summarize(
    plays          = n(),
    share          = n()/nrow(pbp_g1),
    epa_per_play   = mean(epa, na.rm = TRUE),
    success_rate   = mean(success, na.rm = TRUE),
    yards_per_play = mean(yards_gained, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(posteam, desc(plays))
cat("\n--- OFFENSIVE MIX & EFFICIENCY ---\n"); print(off_mix, n = 50)
```

### Regarding the EPA, we can now take a look at the distribution, mainly on runs or passes, and compare the EPA of the Colts to the EPA of the Dolphins

```{r}
epa_dist <- pbp_g1 %>%
  filter(!is.na(posteam), play_type %in% c("run","pass")) %>%
  mutate(is_colts = if_else(posteam == "IND", "Colts", "Opponent"))
ggplot(epa_dist, aes(x = epa, fill = is_colts)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(data = epa_dist %>% group_by(is_colts) %>% summarize(mu = mean(epa, na.rm = TRUE)),
             aes(xintercept = mu, color = is_colts), linewidth = 0.9) +
  scale_x_continuous(labels = number_format(accuracy = 0.1)) +
  labs(title = "EPA distribution (pass & run plays)",
       x = "EPA per play", y = "Count", fill = "", color = "") +
  theme_minimal()
```

### We can look at the situational success rate for each team, Colts and Dolphins, categorized whether it was 1st - 4th down and how many yards to go there were

```{r}
situ <- pbp_g1 %>%
  filter(!is.na(down), !is.na(ydstogo), play_type %in% c("pass","run")) %>%
  mutate(
    ytg_bucket = case_when(
      ydstogo <= 2 ~ "1-2",
      ydstogo <= 5 ~ "3-5",
      ydstogo <= 10 ~ "6-10",
      TRUE ~ "11+"
    ),
    down_lab = paste0("Down ", down)
  ) %>%
  group_by(posteam, down_lab, ytg_bucket) %>%
  summarize(
    plays = n(),
    success_rate = mean(success, na.rm = TRUE),
    epa_per_play = mean(epa, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(situ %>% filter(posteam %in% c("IND", opp)),
       aes(x = ytg_bucket, y = success_rate, group = posteam, color = posteam)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~down_lab) +
  scale_y_continuous(labels = percent) +
  labs(title = "Success rate by down & yards-to-go",
       x = "Yards to go (bucket)", y = "Success rate", color = "Offense") +
  theme_minimal()
```

### Here, we can look at the EPA for each play and categorize it by field position or field zone

```{r}
fieldpos <- pbp_g1 %>%
  filter(!is.na(yardline_100), play_type %in% c("run","pass")) %>%
  mutate(
    field_zone = case_when(
      yardline_100 >= 80 ~ "Own 20-0",
      yardline_100 >= 50 ~ "Own 49-21",
      yardline_100 >= 21 ~ "Opp 49-21",
      TRUE ~ "Red Zone (<=20)"
    )
  ) %>%
  group_by(posteam, field_zone) %>%
  summarize(
    plays = n(),
    epa_per_play = mean(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(fieldpos %>% filter(posteam %in% c("IND", opp)),
       aes(x = field_zone, y = epa_per_play, fill = posteam)) +
  geom_col(position = position_dodge(width = 0.8)) +
  coord_flip() +
  labs(title = "EPA/play by field zone",
       x = "", y = "EPA per play", fill = "Offense") +
  theme_minimal()
```

### This will give us a summary of drives

```{r}
drive_sum <- pbp_g1 %>%
  filter(!is.na(drive), !is.na(posteam)) %>%
  group_by(posteam, drive) %>%
  summarize(
    plays        = n(),
    yards        = sum(yards_gained, na.rm = TRUE),
    points       = sum(points_scored, na.rm = TRUE),   # <-- FIXED
    epa          = sum(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    ended_score  = any(touchdown == 1 | field_goal_result == "made" | safety == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(posteam, desc(points), desc(yards))
```

### This section looks at targets in the game for the Colts and measures their total EPA and number of targets in the game, getting a sense of player involvement in game 1

```{r}
colts_targets <- pbp_g1 %>%
  filter(posteam == team, pass == 1) %>%
  count(receiver_player_name, sort = TRUE, name = "targets")

colts_rushers <- pbp_g1 %>%
  filter(posteam == team, rush == 1) %>%
  count(rusher_player_name, sort = TRUE, name = "rush_att")

epa_by_receiver <- pbp_g1 %>%
  filter(posteam == team, pass == 1, !is.na(receiver_player_name)) %>%
  group_by(receiver_player_name) %>%
  summarize(
    targets = n(),
    epa_sum = sum(epa, na.rm = TRUE),
    epa_mean = mean(epa, na.rm = TRUE),
    .groups = "drop"
  ) %>% arrange(desc(epa_sum))

ggplot(epa_by_receiver, aes(x = targets, y = epa_sum, label = receiver_player_name)) +
  geom_point(size = 3) +
  geom_text_repel(min.segment.length = 0) +
  labs(title = "Colts receivers: total EPA vs targets",
       x = "Targets", y = "Total EPA") +
  theme_minimal()
```

### Penalties and Special Teams

```{r}
penalties <- pbp_g1 %>%
  filter(!is.na(penalty)) %>%
  mutate(off_def = if_else(penalty_team == posteam, "Offense", "Defense")) %>%
  count(penalty_team, off_def, penalty_type, sort = TRUE)
cat("\n--- PENALTIES ---\n"); print(penalties, n = 30)

st_fg <- pbp_g1 %>%
  filter(play_type == "field_goal") %>%
  count(posteam, field_goal_result)
st_punt <- pbp_g1 %>%
  filter(play_type == "punt") %>%
  summarize(punts = n(), avg_punt_yards = mean(yards_gained, na.rm = TRUE))
cat("\n--- SPECIAL TEAMS ---\n"); print(list(field_goals = st_fg, punts = st_punt))
```

### Time-Based Trends - This measures the EPA per play over the course of the game, divided into 4 quarters, for each team, the Colts and the Dolphins

```{r}
epa_by_q <- pbp_g1 %>%
  filter(qtr %in% 1:4, play_type %in% c("pass","run")) %>%
  group_by(posteam, qtr) %>%
  summarize(
    epa_per_play = mean(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(epa_by_q, aes(x = factor(qtr), y = epa_per_play, group = posteam, color = posteam)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "EPA/play by quarter", x = "Quarter", y = "EPA/play", color = "Offense") +
  theme_minimal()
```

### Rolling Pass Rate

```{r}
colts_seq <- pbp_g1 %>%
  filter(posteam == team, play_type %in% c("pass","run")) %>%
  mutate(play_index = row_number(),
         pass_flag  = if_else(play_type == "pass", 1, 0)) %>%
  arrange(play_index)
if (nrow(colts_seq) >= 10) {
  colts_seq_roll <- colts_seq %>%
    mutate(pass_rate_10 = zoo::rollapply(pass_flag, width = 10, FUN = mean,
                                         align = "right", fill = NA_real_))
  ggplot(colts_seq_roll, aes(x = play_index, y = pass_rate_10)) +
    geom_line(linewidth = 1) +
    scale_y_continuous(labels = percent) +
    labs(title = "Colts rolling pass rate (10-play window)",
         x = "Offensive play index", y = "Pass rate") +
    theme_minimal()
}
```

### Compact Console Summary

```{r}
colts_off <- pbp_g1 %>%
  filter(posteam == team) %>%
  summarize(
    plays        = n(),
    pass_plays   = sum(play_type == "pass", na.rm = TRUE),
    rush_plays   = sum(play_type == "run",  na.rm = TRUE),
    total_epa    = sum(epa, na.rm = TRUE),
    mean_epa     = mean(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE)
  )

plays_by_team_type <- pbp_g1 %>%
  filter(!is.na(posteam)) %>%
  count(posteam, play_type, sort = TRUE)

cat("\n--- SUMMARY ---\n")
print(list(
  game_header = tibble(
    season = season, week = week, gameday = gameday,
    home = home_tm, away = away_tm, game_id = game_id, opp_for_colts = opp
  ),
  pbp_rows          = nrow(pbp_g1),
  roster_rows       = nrow(rw_g1),
  officials_rows    = nrow(officials_g1),
  tempo_sec_per_play= tempo,
  plays_by_team_type= plays_by_team_type,
  colts_off_summary = colts_off
))
```

